\documentclass{src/template/phdthesis}
\usepackage{mylegend}

% Hyphenation helpers
\hyphenation{off-load off-load-ing back-pres-sure}

\input{src/acronyms}
\input{src/symbols}
\makeglossaries

\input{src/body/platform_fig}

% some thoughts
\addthought{
	The universe is inherently parallel, and laws of nature are applied everywhere without computational effort and error margin.
	Why is it that hard for a computer in the same universe to do a universe-compatible N-body simulation?
}
\addthought{
	What is the point of reasoning about performance of a program, when that program still contains bugs?
	Can a C program really be considered efficient, if nobody knows whether it is bug-free?
}
\addthought{
	What is `smart' about a smartphone when it lacks the fundamental property of intelligence, namely an understanding of the environment?
	It never knows what I want; I always have to tell it what to do next\ldots
}
\addthought{
	Can a computer make errors?
	It only behaves according to the physical arrangement of matter, which is subject to the laws of nature.
	It is just that a `broken' computer does not live up to the expectations of the user\ldots
}
\addthought{
	Concurrency cannot be invented by a programmer; it only makes his job harder.
}
%\infinitethought{
%	more thoughts
%}

\index{loop!endless|see{endless loop}}
\index{endless loop|see{loop, endless}}
\index{recursion|seealso{recursion}}
\index{easter eggs|see{somewhere}}

\begin{document}

\frontmatter

\selectlanguage{english}
\include{src/front/colophon}
\selectlanguage{dutch}
\include{src/front/titlepage}
%\include{src/front/dedication}

\selectlanguage{english}
\resetacr
\includechapter{src/front/abstract}
\selectlanguage{dutch}
\resetacr
\includechapter{src/front/samenvatting}
\includechapter{src/front/acknowledgment}

\selectlanguage{english}
\formattedTOC
%\formattedLOF
%\formattedLOT

\mainmatter
\resetacr

\includechapter{src/body/1_introduction}
\includechapter{src/body/2_starburst}
\includechapter{src/body/3_programming_model}
\includechapter{src/body/4_hardware}
\includechapter{src/body/5_memory}
\includechapter{src/body/6_concurrency}
\includechapter{src/body/7_conclusion}

\appendix
%\includechapter{src/body/date2012_ideas}
%\includechapter{src/body/atomic_once}
\includechapter{src/body/etymology}

\backmatter
\listofacronyms
\listofreferences
\listofpublications
\formattedindex%
\listoftodos%
\ifdiscussion%
	\include{src/back/colortest}%
\fi%

\includestellingen%
\end{document}
